---
title: "Data 624 Project 2"
author: "Mary Anna Kivenson"
date: "May 3, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(caret)
library(RANN)
```

## Data

```{r}
train <- read.csv("https://raw.githubusercontent.com/mkivenson/Predictive-Analytics/master/Project%202/Datasets/student_train.csv")
test <- read.csv("https://raw.githubusercontent.com/mkivenson/Predictive-Analytics/master/Project%202/Datasets/student_test.csv")
```



```{r}
head(train)
```


#### Summary

First, we take a look at a summary of the data. 

- There are missing values in all of the predictors except for `Pressure.Vacuum` and `Air.Pressurer`
- The target variabale, `PH`, has four missing values that should be removed from the training data
- All predictors except for `Brand.Code` are numeric - `Brand.Code` will be encoded into dummy variables

```{r}
summary(train)
train <- drop_na(train, "PH") #remove rows without a PH value
```





### Preprocessing

#### Split

```{r}
X_train <- subset(train, select = -c(PH))
y_train <- train$PH
X_test <- subset(test, select = -c(PH))
y_test <- test$PH
```


#### Encoding

```{r}
X_train$Brand.A <- ifelse(X_train$Brand.Code == 'A', 1, 0)
X_train$Brand.B <- ifelse(X_train$Brand.Code == 'B', 1, 0)
X_train$Brand.C <- ifelse(X_train$Brand.Code == 'C', 1, 0)
X_train$Brand.D <- ifelse(X_train$Brand.Code == 'D', 1, 0)
X_train <- subset(X_train, select = -c(Brand.Code))
head(X_train)
```


```{r}
aggr(train[,sapply(train, is.numeric)], col=c('navyblue','red'), numbers=TRUE, sortVars=TRUE, labels=names(train), cex.axis=.7, gap=3, ylab=c("Histogram of missing data","Pattern"))
```



```{r}
result <- preProcess(X_train, method = c("knnImpute"), k = 10)
X_train <- predict(result, X_train)
head(X_train)
```

```{r}
library(corrplot)
corrplot(cor(result$data, use = "complete.obs"), method="color", type="lower", tl.col = "black", tl.srt = 5)
```



### PCA

```{r}
rlmPCA <- train(X_train, y_train,
                method = "rlm",
                preProcess = "pca",
                trControl = trainControl(method = "cv"))
rlmPCA 
```




### PLS


```{r}
plsTune <- train(X_train, y_train, 
                 method = "pls", 
                 tuneLength = 20,
                 trControl = trainControl(method = "cv"),
                 preProc = c("center", "scale"))
plsTune
```


```{r}
library(mlbench)

ridgeGrid <- data.frame(.lambda = seq(0, .1, length = 15))
ridgeRegFit <- train(X_train, y_train,
                     method = "ridge", 
                     tuneGrid = ridgeGrid, 
                     trControl = trainControl(method = "cv"),
                     preProc = c("center", "scale"))
ridgeRegFit 
```

```{r}
svmRTuned  <- train(X_train, y_train,
                 method = "svmRadial",
                 #preProc = c("center", "scale"), 
                 tuneLength = 14,
                 trControl = trainControl(method = "cv"))
svmRTuned
```



