---
title: "Assignment 8"
author: "Mary Anna Kivenson"
date: "April 11, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r message=FALSE, warning=FALSE}
library(caret)
library(pls)
library(tidyverse)
library(AppliedPredictiveModeling)
library(corrplot)
library(nnet)
```


# Assignment 8 - Non-Linear Regression Models

## Question 7.2

Friedman (1991) introduced several benchmark data sets create by simulation. One of these simulations used the following nonlinear equation to create data: 

$$y = 10sin(πx_1x_2) + 20(x_3 − 0.5)^2 + 10x_4 +5x_5 + N(0,σ^2)$$

where the x values are random variables uniformly distributed between [0, 1] (there are also 5 other non-informative variables also created in the simulation). The package mlbench contains a function called `mlbench.friedman1` that simulates these data:


```{r}
library(mlbench) 
set.seed(200) 
trainingData <- mlbench.friedman1(200, sd = 1) 
  ## We convert the 'x' data from a matrix to a data frame  
## One reason is that this will give the columns names. 
trainingData$x <- data.frame(trainingData$x) 
## Look at the data using 
featurePlot(trainingData$x, trainingData$y) 
## or other methods. 
## This creates a list with a vector 'y' and a matrix 
## of predictors 'x'. Also simulate a large test set to 
## estimate the true error rate with good precision: 
testData <- mlbench.friedman1(5000, sd = 1)
testData$x <- data.frame(testData$x) 
```

Tune several models on these data. For example:

```{r}
library(caret)
knnModel <- train(x = trainingData$x, 
                  y = trainingData$y, 
                  method = "knn", 
                  preProc = c("center", "scale"), 
                  tuneLength = 10) 

knnModel
```


```{r}
knnPred <- predict(knnModel, newdata = testData$x)
postResample(pred = knnPred, obs = testData$y)
```


#### Neural Network Model


```{r eval=TRUE, message=FALSE, warning=FALSE, include=FALSE}
nnetFit <- avNNet(x = trainingData$x, 
                y = trainingData$y, 
                size = 5, 
                decay = 0.01, 
                linout = TRUE, 
                preProc = c("center", "scale"),
                maxit = 500,
                MaxNWts=5*(ncol(trainingData$x) + 1) + 5 + 1)
```

```{r}
nnetFit
```

```{r}
nnetPred <- predict(nnetFit, newdata = testData$x)
postResample(pred = nnetPred, obs = testData$y)
```




#### Multivariate Adaptive Regression Splines

```{r message=FALSE, warning=FALSE}
library(earth)
marsGrid <- expand.grid(.degree = 1:2, .nprune = 2:38)
marsTuned <- train(x = trainingData$x, 
                   y = trainingData$y,
                   method = "earth",
                   tuneGrid = marsGrid,
                   trControl = trainControl(method = "cv"))
summary(marsTuned)
```


```{r}
marsPred <- predict(marsTuned, newdata = testData$x)
postResample(pred = marsPred, obs = testData$y)
```


#### Support Vector Machines

```{r message=FALSE, warning=FALSE}
library(kernlab)
svmRTuned <- train(x = trainingData$x, 
                   y = trainingData$y,
                   method = "svmRadial", 
                   preProc = c("center", "scale"), 
                   tuneLength = 14, 
                   trControl = trainControl(method = "cv"))
svmRTuned
```


```{r}
svmPred <- predict(svmRTuned, newdata = testData$x)
postResample(pred = svmPred, obs = testData$y)
```


The Multivariate Adaptive Regression Splines model produces the lowest RMSE, highest Rsquared, and lowest MAE. The MARS model selected 5 out of 10 predictors (X1-X5); these are all the informative predictors. 


## Question 7.5

Exercise 6.3 describes data for a chemical manufacturing process. Use the same data imputation, data splitting, and pre-processing steps as before and train several nonlinear regression models.

```{r}
data(ChemicalManufacturingProcess)
cmp <- ChemicalManufacturingProcess[, -nearZeroVar(ChemicalManufacturingProcess)]

#kNN imputation will be used to fill the missing values in the dataset.
cmp <- preProcess(as.data.frame(cmp), method = "knnImpute", k = 10)$data

# test train split
set.seed(0)
smp_size <- floor(0.8 * nrow(cmp))
train_ind <- sample(seq_len(nrow(cmp)), size = smp_size)

Xtrain <- cmp[train_ind, -1]
Xtest <- cmp[-train_ind, -1]

ytrain <- cmp[train_ind, 1]
ytest <- cmp[-train_ind, 1]
```

(a) Which nonlinear regression model gives the optimal resampling and test set performance? 

#### K Nearest Neighbor

```{r}
knnModel <- train(x = Xtrain, 
                  y = ytrain, 
                  method = "knn", 
                  preProc = c("center", "scale"), 
                  tuneLength = 10) 

knnModel
```


```{r}
knnPred <- predict(knnModel, newdata = Xtest)
postResample(pred = knnPred, obs = ytest)
```


#### Neural Network Model


```{r eval=TRUE, message=FALSE, warning=FALSE, include=FALSE}
nnetFit <- avNNet(x = Xtrain, 
                y = ytrain, 
                size = 5, 
                decay = 0.01, 
                linout = TRUE, 
                preProc = c("center", "scale"),
                maxit = 500,
                MaxNWts=5*(ncol(Xtrain) + 1) + 5 + 1)
```

```{r}
nnetFit
```

```{r}
nnetPred <- predict(nnetFit, newdata = Xtest)
postResample(pred = nnetPred, obs = ytest)
```


#### Multivariate Adaptive Regression Splines

```{r message=FALSE, warning=FALSE}
library(earth)
marsGrid <- expand.grid(.degree = 1:2, .nprune = 2:38)
marsTuned <- train(x = Xtrain, 
                   y = ytrain,
                   method = "earth",
                   tuneGrid = marsGrid,
                   trControl = trainControl(method = "cv"))
summary(marsTuned)
```


```{r}
marsPred <- predict(marsTuned, newdata = Xtest)
postResample(pred = marsPred, obs = ytest)
```


#### Support Vector Machines

```{r message=FALSE, warning=FALSE}
library(kernlab)
svmRTuned <- train(x = Xtrain, 
                   y = ytrain,
                   method = "svmRadial", 
                   preProc = c("center", "scale"), 
                   tuneLength = 14, 
                   trControl = trainControl(method = "cv"))
svmRTuned
```


```{r}
svmPred <- predict(svmRTuned, newdata = Xtest)
postResample(pred = svmPred, obs = ytest)
```

The SVM model produces the best results when resampling on the test set, which a much higher R-squared and lower RMSE than the other models.

(b) Which predictors are most important in the optimal nonlinear regression model? Do either the biological or process variables dominate the list? How do the top ten important predictors compare to the top ten predictors from the optimal linear model? 

For the SVM model, which is the optimal nonlinear regression model, the following chart shows the ten most important predictors. Neither manufacturing nor biological variables dominate the list completely, though manufacturing processes make up the top four predictors. This is similar to the predictor importance results of the optimal linear model, which includes all the same variables other than `BiologicalMaterial12`. The order of importance is also different, but both models have the same manufacturing processes in the top 4.

```{r}
plsTune <- train(Xtrain, 
                 ytrain,
                 method = "pls",
                 tuneLength = 30,
                 preProc = c("center", "scale"),
                 trControl =  trainControl(method = 'cv', 10))

plot(varImp(svmRTuned), top = 10, main = "SVM")
plot(varImp(plsTune), top = 10, main = "Linear")

```


(c) Explore the relationships between the top predictors and the response for the predictors that are unique to the optimal nonlinear regression model. Do these plots reveal intuition about the biological or process predictors and their relationship with yield?

The only predictor that is unique to the SVM model is `BiologicalMaterial12` - the relationship between this predictor and yield is linear, positive, and somewhat weak. 

```{r}
plot(Xtrain$BiologicalMaterial12, ytrain)
```

