---
title: "Market Basket Analysis"
author: "Mary Anna Kivenson"
date: "May 3, 2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Data

Imagine 10000 receipts sitting on your table. Each receipt represents a transaction with items that were purchased. The receipt is a representation of stuff that went into a customer’s basket - and therefore ‘Market Basket Analysis’. That is exactly what the Groceries Data Set contains: a collection of receipts with each line representing 1 receipt and the items purchased. Each line is called a transaction and each column in a row represents an item. The data set is attached.

Your assignment is to use R to mine the data for association rules. You should report support, confidence and lift and your top 10 rules by lift. 

```{r message=FALSE, warning=FALSE}
library(plyr)
library(tidyr)
library(readxl)
library(knitr)
library(ggplot2)
library(lubridate)
library(arules)
library(arulesViz)
```



#### Load Data

```{r}
df <- read.transactions("https://raw.githubusercontent.com/mkivenson/Predictive-Analytics/master/Market%20Basket%20Analysis/GroceryDataSet.csv", header = FALSE, sep = ",")
```





#### Use `arules` to return top 10 rules by lift

```{r eval=TRUE, message=TRUE, warning=TRUE}
rules <- apriori(df, parameter = list(supp = 0.001, conf = 0.8))
```


```{r}
inspect(rules[0:10])
```
